humanCdGF_predict_cfu

### 2016-11-21 code/HumanCdGF_predict_cfu.Rmd
add om_loo to colonization rate
also updated plots with unidentifiable title

#### merge OOB and leave 1 cage out dataframes
all_rf <- rbind(cbind(rf_prediction, data = rep('OOB', nrow(rf_prediction))),
                cbind(LOcO_probs, data = rep('LOO', nrow(LOcO_probs))),
                cbind(om_pred, data = rep('1_mouse', nrow(om_pred))))

ggplot(data = all_rf, aes(x=observed, y=predicted, color = data)) + 
  geom_abline(color = 'gray', size = 0.5, linetype = 2, slope = 1) + 
  # geom_point() + 
  geom_smooth(method = 'lm', se = F) +
  stat_summary(fun.y=median, geom="point", aes(group = data)) + 
  stat_summary(fun.data = 'median_hilow') +
  xlim(c(5,9)) + ylim(c(5,9)) +
  labs(x= 'Observed Log10 CFU', y = 'Predicted Log10 CFU',
       title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
  theme_bw() + 
  scale_color_discrete(name = 'Model',
                       breaks = c('OOB', 'LOO', '1_mouse'),
                       labels = c(bquote('OOB (R'^'2'*' = '*.(round(tail(rf_n_features$rsq, 1), 2))*')'),
                                  bquote('LOO (R'^'2'*' = '*.(round(LOcO_rsq, 2))*')'),
                                  bquote('OM (R'^'2'*' = '*.(round(om_rsq,2))*')'))) + 
  theme(legend.justification=c(0,1), legend.position=c(0,1))

@@

layout(matrix(c(1:9), ncol=3, byrow=TRUE))
for(otu in seq_along(importance_om$tax)){
    partialPlot(rf_n_features, data.frame(cfu_shared_colonized.df), x.var=rownames(importance_om)[otu],
                xlab='', main=paste(importance_om$tax_label[otu]))
    par(new = T)
    plot(y = partial_df[ , 'log_cfu_med'],
         x = partial_df[ , rownames(importance_om)[otu]], 
         pch = NA, bg = 'gray', xlab = NA, ylab = NA, axes = F)
    axis(side = 4)
    mtext(side = 4, line = 3, 'Log10 CFU')
    if(length(features_cor[[otu]]) > 1){
    mtext(side = 1, line = 3, cex = 0.6,
          paste('Bacteria removed in preprocessing due to correlation to this OTU:\n',
                paste(get_tax(row_list = features_cor[[otu]])$tax_label, collapse = ',')))
    }
# To color points by mouse - Change pch to 21 and un-comment following lines
  lapply(seq_along(unique(partial_df$cage_id)), function(m){
    mouse <- unique(partial_df$cage_id)[m]
    points(y = partial_df[partial_df$cage_id %in% mouse, 'log_cfu_med'],
           x = partial_df[partial_df$cage_id %in% mouse , rownames(importance_om)[otu]], 
           pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
  })
}
```

@@

#### merge OOB and leave 1 cage out dataframes
all_cr_rf <- rbind(cbind(cr_prediction, data = rep('OOB', nrow(cr_prediction))),
                cbind(LOcO_cr_probs, data = rep('LOO', nrow(LOcO_cr_probs))),
                cbind(om_pred_cr, data = rep('1_mouse', nrow(om_pred_cr))))

ggplot(data = all_cr_rf, aes(x=observed, y=predicted, color = data)) + 
  geom_abline(color = 'gray', size = 0.5, linetype = 2, slope = 1) + 
  # geom_point() + 
  geom_smooth(method = 'lm', se = F) +
  stat_summary(fun.y=median, geom="point", aes(group = data)) + 
  stat_summary(fun.data = 'median_hilow') +
  xlim(c(min(all_cr_rf[, c('observed','predicted')]),
         max(all_cr_rf[, c('observed','predicted')]))) + 
  ylim(c(min(all_cr_rf[, c('observed','predicted')]),
         max(all_cr_rf[, c('observed','predicted')]))) +
  labs(x= 'Observed Log10 CFU', y = 'Predicted Log10 CFU',
       title = "Random Forest Prediction of C. Difficile Colonization Rate from Day 0 Community") + 
  theme_bw() + 
  scale_color_discrete(name = 'Model',
                       breaks = c('OOB', 'LOO', '1_mouse'),
                       labels = c(bquote('OOB (R'^'2'*' = '*.(round(tail(rf_col_rate$rsq, 1), 2))*')'),
                                  bquote('LOO (R'^'2'*' = '*.(round(LOcO_rsq_cr, 2))*')'),
                                  bquote('OM (R'^'2'*' = '*.(round(om_rsq_cr,2))*')'))) + 
  theme(legend.justification=c(0,1), legend.position=c(0,1))

@@

cage_color <- rainbow(length(unique(cr_prediction$cage_id)))
for (otu in seq_along(cr_features)) {
    partialPlot(rf_col_rate, shared_cr_df, cr_features[otu],
                xlab='', main=paste(get_tax(row_list = cr_features)$tax_label[otu]))#, log = 'x')
    par(new = T)
    plot(y = shared_col_rate_df[ , 'ratio'],
         x = shared_col_rate_df[ , cr_features[otu]], 
         pch = NA, bg = 'gray', xlab = NA, ylab = NA, axes = F)
    axis(side = 4)
    mtext(side = 4, line = 3, 'Relative Colonization')
    if(length(cr_features_cor[[otu]]) > 1){
      mtext(side = 1, line = 3, cex = 0.6,
            paste('Bacteria removed in preprocessing due to correlation to this OTU:\n',
                  paste(get_tax(row_list = cr_features_cor[[otu]])$tax_label, collapse = ',')))
    }
# To color points by mouse - Change pch to 21 and un-comment following lines
  lapply(seq_along(unique(shared_col_rate_df$cage_id)), function(m){
    mouse <- unique(shared_col_rate_df$cage_id)[m]
    points(y = shared_col_rate_df[shared_col_rate_df$cage_id %in% mouse, 'ratio'],
           x = shared_col_rate_df[shared_col_rate_df$cage_id %in% mouse , cr_features[otu]], 
           pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
  })
}
```

>>>>

# one mouse leave one out
om_loo_pred <- c()
for(i in 1:iters){
    set.seed(i)
    one_mouse <- one_mouse_df %>% 
      group_by(cage_id) %>% 
      sample_n(1) %>% ungroup() 
    for (n in unique(as.character(one_mouse$cage_id))){
      train <- filter(one_mouse, !cage_id == n)
      test <- one_mouse_df %>% 
        filter(cage_id == n)
      set.seed(seed)
      rf_om_loo <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
                                ntree = 500)
      om_loo_pred <- rbind(om_loo_pred,
                           data.frame(select(test, mouse_id, cage_id, log_cfu_med),
                           predicted=predict(rf_om_loo, test)))
    }
}

om_loo_rsq <- 1 - (sum((om_loo_pred$log_cfu_med - om_loo_pred$predicted)^2)/
               sum((om_loo_pred$log_cfu_med - mean(om_loo_pred$log_cfu_med))^2))
om_loo_pred <- rename(om_loo_pred, observed = log_cfu_med)

#### merge OOB and leave 1 cage out dataframes
all_rf <- rbind(cbind(rf_prediction, data = rep('OOB', nrow(rf_prediction))),
                cbind(LOcO_probs, data = rep('LOO', nrow(LOcO_probs))),
                cbind(om_loo_pred, data = rep('om_loo', nrow(LOcO_probs))),
                cbind(om_pred, data = rep('1_mouse', nrow(om_pred))))

ggplot(data = all_rf, aes(x=observed, y=predicted, color = data)) + 
  #geom_abline(color = 'gray', size = 0.5, linetype = 2, slope = 1) + 
  # geom_point() + 
  #geom_smooth(method = 'lm', se = F) +
  stat_summary(fun.y=median, geom="point", aes(group = data)) + 
  stat_summary(fun.data = 'median_hilow') +
  xlim(c(5,9)) + ylim(c(5,9)) +
  labs(x= 'Observed Log10 CFU', y = 'Predicted Log10 CFU',
       title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
  theme_bw() + 
  scale_color_discrete(name = NULL,
                       breaks = c('OOB', 'LOO', '1_mouse', 'om_loo'),
                       labels = c(bquote('OOB (R'^'2'*' = '*.(round(tail(rf_n_features$rsq, 1), 2))*')'),
                                  bquote('LOO (R'^'2'*' = '*.(round(LOcO_rsq, 2))*')'),
                                  bquote('OM (R'^'2'*' = '*.(round(om_rsq,2))*')'),
                                  bquote('OM-LOO (R'^'2'*' = '*.(round(om_loo_rsq,2))*')'))) + 
  theme(legend.justification=c(0,1), legend.position=c(0,1))

@@

layout(matrix(c(1:9), ncol=3, byrow=TRUE))
for(otu in seq_along(importance_om$tax)){
    partialPlot(rf_n_features, data.frame(cfu_shared_colonized.df), x.var=rownames(importance_om)[otu],
                xlab='', main = NULL)
    if(length(features_cor[[otu]]) > 1){
      title(main = paste0('Unidentifiable Feature (', 
                          get_tax(row_list = rownames(importance_om)[otu])$tax_label,
                          ' Shown)'))
      mtext(side = 1, line = 3, cex = 0.6,
        paste('Correlated features: \n', 
              paste(get_tax(row_list = features_cor[[otu]])$tax_label, collapse = ', ')))
      } else {
      title(paste(get_tax(row_list = rownames(importance_om)[otu])$tax_label))
        }
    par(new = T)
    plot(y = partial_df[ , 'log_cfu_med'],
         x = partial_df[ , rownames(importance_om)[otu]], 
         pch = NA, bg = 'gray', xlab = NA, ylab = NA, axes = F)
    axis(side = 4)
    mtext(side = 4, line = 3, 'Log10 CFU')
# To color points by mouse - Change pch to 21 and un-comment following lines
  lapply(seq_along(unique(partial_df$cage_id)), function(m){
    mouse <- unique(partial_df$cage_id)[m]
    points(y = partial_df[partial_df$cage_id %in% mouse, 'log_cfu_med'],
           x = partial_df[partial_df$cage_id %in% mouse , rownames(importance_om)[otu]], 
           pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
  })
}
```

@@

# one mouse leave one out
om_loo_pred_cr <- c()
for(i in 1:iters){
    set.seed(i)
    one_mouse <- om_cr_df %>% 
      group_by(cage_id) %>% 
      sample_n(1) %>% ungroup() 
    for (n in unique(as.character(one_mouse$cage_id))){
      train <- filter(one_mouse, !cage_id == n)
      test <- om_cr_df %>% 
        filter(cage_id == n)
      set.seed(seed)
      rf_om_loo_cr <- randomForest(ratio ~ ., data = select(train, -mouse_id, -cage_id),
                                ntree = 500)
      om_loo_pred_cr <- rbind(om_loo_pred_cr,
                           data.frame(select(test, mouse_id, cage_id, ratio),
                           predicted=predict(rf_om_loo_cr, test)))
    }
}

om_loo_rsq_cr <- 1 - (sum((om_loo_pred_cr$ratio - om_loo_pred_cr$predicted)^2)/
               sum((om_loo_pred_cr$ratio - mean(om_loo_pred_cr$ratio))^2))
om_loo_pred_cr <- rename(om_loo_pred_cr, observed = ratio)

#### merge OOB and leave 1 cage out dataframes
all_cr_rf <- rbind(cbind(cr_prediction, data = rep('OOB', nrow(cr_prediction))),
                cbind(LOcO_cr_probs, data = rep('LOO', nrow(LOcO_cr_probs))),
                cbind(om_loo_pred_cr, data = rep('OM_LOO', nrow(om_loo_pred_cr))),
                cbind(om_pred_cr, data = rep('1_mouse', nrow(om_pred_cr))))

ggplot(data = all_cr_rf, aes(x=observed, y=predicted, color = data)) + 
  #geom_abline(color = 'gray', size = 0.5, linetype = 2, slope = 1) + 
  # geom_point() + 
  #geom_smooth(method = 'lm', se = F) +
  stat_summary(fun.y=median, geom="point", aes(group = data)) + 
  stat_summary(fun.data = 'median_hilow') +
  xlim(c(min(all_cr_rf[, c('observed','predicted')]),
         max(all_cr_rf[, c('observed','predicted')]))) + 
  ylim(c(min(all_cr_rf[, c('observed','predicted')]),
         max(all_cr_rf[, c('observed','predicted')]))) +
  labs(x= 'Observed Log10 CFU', y = 'Predicted Log10 CFU',
       title = "Random Forest Prediction of C. Difficile Colonization Rate from Day 0 Community") + 
  theme_bw() + 
  scale_color_discrete(name = NULL,
                       breaks = c('OOB', 'LOO', '1_mouse', 'OM_LOO'),
                       labels = c(bquote('OOB (R'^'2'*' = '*.(round(tail(rf_col_rate$rsq, 1), 2))*')'),
                                  bquote('LOO (R'^'2'*' = '*.(round(LOcO_rsq_cr, 2))*')'),
                                  bquote('OM (R'^'2'*' = '*.(round(om_rsq_cr,2))*')'),
                                  bquote('OM_LOO (R'^'2'*' = '*.(round(om_loo_rsq_cr,2))*')'))) + 
  theme(legend.justification=c(0,1), legend.position=c(0,1))

@@

cage_color <- rainbow(length(unique(cr_prediction$cage_id)))
for (otu in seq_along(cr_features)) {
    partialPlot(rf_col_rate, shared_cr_df, cr_features[otu],
                xlab='', main = NULL)
    if(length(cr_features_cor[[otu]]) > 1){
      title(main = paste0('Unidentifiable Feature (', 
                          get_tax(row_list = cr_features[otu])$tax_label,
                          ' Shown)'))
      mtext(side = 1, line = 3, cex = 0.6,
        paste('Correlated features: \n', 
              paste(get_tax(row_list = cr_features_cor[[otu]])$tax_label, collapse = ', ')))
      } else {
      title(paste(get_tax(row_list = cr_features[otu])$tax_label))
        }
  par(new = T)
    plot(y = shared_col_rate_df[ , 'ratio'],
         x = shared_col_rate_df[ , cr_features[otu]], 
         pch = NA, bg = 'gray', xlab = NA, ylab = NA, axes = F)
    axis(side = 4)
    mtext(side = 4, line = 3, 'Relative Colonization')
# To color points by mouse - Change pch to 21 and un-comment following lines
  lapply(seq_along(unique(shared_col_rate_df$cage_id)), function(m){
    mouse <- unique(shared_col_rate_df$cage_id)[m]
    points(y = shared_col_rate_df[shared_col_rate_df$cage_id %in% mouse, 'ratio'],
           x = shared_col_rate_df[shared_col_rate_df$cage_id %in% mouse , cr_features[otu]], 
           pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
  })
}
```
